Arrows for Parallel Computation

### 2018-01-07 paper/content/benchmarks.tex
headings consistent

\subsection{Measurement Platform}
We start by explaining the hardware and software stack and outline the benchmark programs and motivation for choosing them. We also shortly address hyper-threading and why we do not use it in our benchmarks.

\subsubsection{Hardware and Software}

@@

\subsubsection{Distributed Memory}

>>>>

\subsection{Measurement platform}
We start by explaining the hardware and software stack and outline the benchmark programs and motivation for choosing them. We also shortly address hyper-threading and why we do not use it in our benchmarks.

\subsubsection{Hardware and software}

@@

\subsubsection{Distributed memory}

### 2018-01-07 paper/content/parrows.tex
headings consistent

\section{Parallel Arrows}
\label{sec:parallel-arrows}
While Arrows are a general interface to computation, we introduce here specialised Arrows as a general interface to \textit{parallel computations}. We present the |ArrowParallel| type class and explain the reasoning behind it before discussing some parallel Haskell implementations and basic extensions.
\subsection{The ArrowParallel type class}
A~parallel computation (on functions) can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Section~\ref{sec:parEvalNIntro}).
Translating this into Arrow terms gives us a new operator |parEvalN| that lifts a list of Arrows |[arr a b]| to a parallel Arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to |evalN| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{code}
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of |parEvalN|, parallel execution is yet another Arrow combinator. But as the implementation may differ depending on the actual type of the Arrow |arr| - or even the input |a| and output |b| - and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{code}
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we introduce an additional |conf| parameter as we do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|. % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% % \begin{figure}[h]
% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{code}
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
By restricting the implementations of our backends to a specific |conf| type, we also get interoperability between backends for free. We can parallelize one part of a program using one backend, and parallelize the next with another one.
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
% , as we will see in the implementation of the GpH Haskell and the |Par| Monad backends.

\subsection{ArrowParallel instances}
With the type class defined, we will now give implementations of it with GpH, the |Par| Monad and Eden.
\subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
The GpH implementation of |ArrowParallel| is implemented in a straightforward manner in Fig.~\ref{fig:ArrowParallelMulticore}, but a bit different compared to the variant from Section \ref{sec:GpHIntro}. We use |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters. Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

>>>>

\section{Parallel Arrows}
\label{sec:parallel-arrows}
While Arrows are a general interface to computation, we introduce here specialised Arrows as a general interface to \textit{parallel computations}. We present the |ArrowParallel| type class and explain the reasoning behind it before discussing some parallel Haskell implementations and basic extensions.
\subsection{The |ArrowParallel| type class}
A~parallel computation (on functions) can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Section~\ref{sec:parEvalNIntro}).
Translating this into Arrow terms gives us a new operator |parEvalN| that lifts a list of Arrows |[arr a b]| to a parallel Arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to |evalN| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{code}
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of |parEvalN|, parallel execution is yet another Arrow combinator. But as the implementation may differ depending on the actual type of the Arrow |arr| - or even the input |a| and output |b| - and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{code}
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we introduce an additional |conf| parameter as we do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|. % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% % \begin{figure}[h]
% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{code}
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
By restricting the implementations of our backends to a specific |conf| type, we also get interoperability between backends for free. We can parallelize one part of a program using one backend, and parallelize the next with another one.
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
% , as we will see in the implementation of the GpH Haskell and the |Par| Monad backends.

\subsection{|ArrowParallel| instances}
With the type class defined, we will now give implementations of it with GpH, the |Par| Monad and Eden.
\subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
The GpH implementation of |ArrowParallel| is implemented in a straightforward manner in Fig.~\ref{fig:ArrowParallelMulticore}, but a bit different compared to the variant from Section \ref{sec:GpHIntro}. We use |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters. Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

### 2018-01-07 paper/content/utilityFunctions.tex
headings consistent

\section{Omitted Function Definitions}
\label{app:omitted}
We have omitted some function definitions in the main text for
brevity, and redeem this here.
%
%
%
We begin with warping Eden's build-in |RemoteData| to |Future| in
Figure~\ref{fig:RDFuture}

@@

matAdd = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
\end{code}
	\caption{|prMMTr|, |splitMatrix| and |matAdd| definition.}
	\label{fig:torus_example_rest}
\end{figure}

\section{Syntactic Sugar} \label{syntacticSugar}
Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.
For basic Arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two Arrows |arr a b| and |arr c d| into an Arrow |arr (a, c) (b, d)| which does both computations at once. This can easily be translated into a parallel version |parstar| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
):
% \begin{figure}[h]
\begin{code}
(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{code}
% \caption{Definition of |parstar|, the parallel version of |***|.}
% \label{fig:par***}
% \end{figure}
% With this we can analogously to the serial |&&&|
We define the parallel |parand| % (Fig.~\ref{fig:par&&&}) 
in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):
% \begin{figure}[h]
\begin{code}
(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{code} % $ %% formatting
% \caption{Definition of |parand| - the parallel version of |&&&|.}
% \label{fig:par&&&}
% \end{figure}

>>>>

\section{Additional function definitions}
\label{app:omitted}
We have omitted some function definitions in the main text for
brevity, and redeem this here.
%
%
%
We begin with warping Eden's build-in |RemoteData| to |Future| in
Figure~\ref{fig:RDFuture}

@@

matAdd = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
\end{code}
	\caption{|prMMTr|, |splitMatrix| and |matAdd| definition.}
	\label{fig:torus_example_rest}
\end{figure} %$

\section{Syntactic sugar} \label{syntacticSugar}
Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.
For basic Arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two Arrows |arr a b| and |arr c d| into an Arrow |arr (a, c) (b, d)| which does both computations at once. This can easily be translated into a parallel version |parstar| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
):
% \begin{figure}[h]
\begin{code}
(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{code}
% \caption{Definition of |parstar|, the parallel version of |***|.}
% \label{fig:par***}
% \end{figure}
% With this we can analogously to the serial |&&&|
We define the parallel |parand| % (Fig.~\ref{fig:par&&&}) 
in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):
% \begin{figure}[h]
\begin{code}
(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{code} % $ %% formatting
% \caption{Definition of |parand| - the parallel version of |&&&|.}
% \label{fig:par&&&}
% \end{figure}
